The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-05 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
finish draft1 review

Bounds on parameters raise some interesting questions about how 
uncertainty in parameter estimates should be computed or reported. That is, the
traditional "standard errors" are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active.

@@

Let us try to fix (mask) the first parameter in the first two example problems.

@@

One of the referees pointed out that nlsr::nlfb() is quite slow on a
"large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
However, we note that problems in large numbers of parameters
will be problematic for Gauss-Newton methods because they create large
computational matrices for the Jacobian or its inner product. The method
"Rcgmin" from package `optimx` (@p-optimx) can be much 
faster, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods.

@@

Indeed, the argument only gets used when `nls()` is about to return its result
object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same 
name could
be confusing.) Despite this, the model frame is used within the function code in 
the form of the object
`mf`. We feel that users could benefit from more extensive documentation and 
examples of its use since it is used to implement features like `subset`.

@@

If `nls()` reaches a point where it cannot continue but has not found a point 
where the relative offset convergence criterion is met, it may simply exit,
especially if a "singular gradient" (singular Jacobian) is found. However, 
this may occur AFTER the function has made considerable progress in reducing 
the sum of squared residuals. 
<!-- An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package. -->
<!-- ?? AB: Are we far enough along to include reference? -->
Here is an abbreviated example:

>>>>

Bounds on parameters raise some interesting questions about how 
uncertainty in parameter estimates should be computed or reported. That is, the
traditional "standard errors" are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active. Transformations to
implicitly enforce bounds avoid this issue, but at the cost of changing the 
form of the model.

@@

Let us try to fix (mask) the first parameter in the first two example problems. We might
want to do this if it is considered that the asymptote is known from external sources,
but we wish later to have it available to be estimated.

@@

One of the referees pointed out that nlsr::nlfb() is quite slow on a
"large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
However, we note that problems in large numbers of parameters
will be problematic for Gauss-Newton methods because they create large
computational matrices for the Jacobian or its inner product. This was
confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
the Penalty 1 problem of various sizes, where the latter, which is 
compiled Fortran internally, was moderately faster. However, the method
"Rcgmin" from package `optimx` (@p-optimx) can be much 
faster, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods.

@@

Indeed, the argument only gets used when `nls()` is about to return its result
object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same 
name could
be confusing.) Despite this, the model frame is used within the function code in 
the form of the object
`mf`. We feel that users could benefit from more extensive documentation and 
examples of its use since it is used to implement features like `subset`. Moreover,
it provides access to resources that may be important in further calculations on
models.

@@

If `nls()` reaches a point where it cannot continue but has not found a point 
where the relative offset convergence criterion is met, it may simply exit,
especially if a "singular gradient" (singular Jacobian) is found. However, 
this may occur **after** considerable progress has been made in reducing 
the sum of squared residuals. 
<!-- An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package. -->
<!-- ?? AB: Are we far enough along to include reference? -->
Here is an abbreviated example:
